---
title: "NYPD Shooting Incident Project"
author: "Shruti Chandrasekaran"
date: "2025-02-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# NYPD Shooting Incident Data Report

This report consists of shooting incident that occurred in City Of NewYork from 2006 through the end of the previous calendar year. This data is obtained from **DATA.GOV**, where dataset is intended for public access and use. 

## Libraries
The library used in this project for analyzing and visualizing data is tidyverse and lubridate. By installing the tidyverse package and loading it using library(tidyverse), I will be able to use dplyr for data manipulation and ggplot2 for visualization. The lubridate package helps with date and time conversions.

```{r}
library(tidyverse)
library(lubridate)
```

## Importing Data

The data is import from the above mentioned site, the csv file consist of details about the shooting occured in the city of new york.

```{r, loading}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_incident <- read.csv(url)

```



```{r}
summary(nypd_incident)
```

```{r}
head(nypd_incident)
```
## Data Cleaning

Once the data is imported we will proceed with the data cleaning process. The summary function provides a quick overview of the key characteristics of the data which will help us with the data cleaning process.

 * The column names and data types for for OCCUR_DATE and OCCUR_TIME is changed to date and time respectively.Additionally, YEAR and MONTH columns are included for further analysis.
 
 * Unnecessary columns are removed, and some columns are converted to factors for better analysis.
 
 * To remove duplicate records

```{r} 
 nypd_incident <- nypd_incident %>%
  rename(DATE = OCCUR_DATE, 
         TIME = OCCUR_TIME) %>%
  mutate(DATE = mdy(DATE),  
         TIME = hms(TIME),
         YEAR = year(DATE),
         MONTHS = month(DATE, label = TRUE))
``` 


```{r}
nypd_incident <- nypd_incident %>%
  select(DATE,TIME,BORO,PRECINCT,STATISTICAL_MURDER_FLAG,
            PERP_AGE_GROUP,PERP_SEX,PERP_RACE, VIC_AGE_GROUP, VIC_SEX,VIC_RACE, YEAR, MONTHS)
``` 


```{r}
nypd_incident$BORO = as.factor(nypd_incident$BORO)
nypd_incident$PERP_AGE_GROUP = as.factor(nypd_incident$PERP_AGE_GROUP)
nypd_incident$PERP_SEX = as.factor(nypd_incident$PERP_SEX)
nypd_incident$PERP_RACE = as.factor(nypd_incident$PERP_RACE)
nypd_incident$VIC_AGE_GROUP = as.factor(nypd_incident$VIC_AGE_GROUP)
nypd_incident$VIC_SEX = as.factor(nypd_incident$VIC_SEX)
nypd_incident$VIC_RACE = as.factor(nypd_incident$VIC_RACE)
```    

```{r}
nypd_incident <- unique(nypd_incident)

```    

```{r}
summary(nypd_incident)
```

## Data Analysis And Visualization

Once the data cleaning is done the next step is to proceed with data analysis and visualization of data.

```{r}

nypd_boro_year <- nypd_incident %>%
  group_by(BORO, YEAR) %>%
  summarize(incident_count = n(), .groups = "drop")  

ggplot(nypd_boro_year, aes(x = YEAR, y = incident_count)) +
  geom_bar(stat = "identity", fill = "lightblue", width =0.3) +
  labs(title = "NYPD Shooting Incident By Year", x = "YEAR", y = "Incident Count") 

```



```{r}

incident_by_months <- nypd_incident %>%
  group_by(MONTHS) %>%      
  summarise(incident_count = n()) %>% 
  ungroup() 

ggplot(incident_by_months, aes(x = MONTHS, y = incident_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Adds color for clarity
  geom_text(aes(label = incident_count), 
            vjust = -0.5, color = "black", size = 3) +
  labs(title = "Number of Incidents by Month", x = "Month", y = "Incident Count") 

```

```{r}
nypd_last_5years <- nypd_incident %>%
  group_by(BORO, YEAR) %>%
  
  summarize(incident_count = n(), .groups = "drop")
  
 ggplot(nypd_last_5years, aes(x = YEAR, y = incident_count, color = BORO)) +
  geom_point(size = 3, alpha = 0.7) + 
  geom_line(aes(group = BORO), linewidth = 1) +  
  labs(title = "Number of Incidents by Borough and Year", 
       x = "Year", 
       y = "Incident Count",
       color = "Borough") +
  theme_minimal()

```
```{r}
incident_by_age <- filter(nypd_incident, VIC_AGE_GROUP!= 1022 )

ggplot(incident_by_age, aes(x = VIC_AGE_GROUP)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Shooting Incidents by Age Group of Perps",
       x = "Age Group",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```  

```{r}

nypd_vic <- incident_by_age %>%
  select(BORO, VIC_AGE_GROUP)
colSums(is.na(nypd_vic))

ggplot(nypd_vic, aes(x = BORO, fill = VIC_AGE_GROUP)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Incidents by Victims Age Group and Borough",
       x = "Borough", y = "Incident Count")
```  


## Data Modelling

Once data analysis and visualization are completed, we will proceed with data modeling, which is a mathematical representation used to identify patterns, relationships, and dependencies. This enables us to make predictions or classifications based on input data. Since Brooklyn had the highest number of incidents, we focus on it for modeling.


```{r}
brooklyn_shootings <- nypd_incident %>% 
filter(BORO=='BROOKLYN')%>% 
  group_by(YEAR) %>%
  summarise(number_of_crimes=n())


mod <- lm(number_of_crimes ~YEAR, brooklyn_shootings)

shooting_pred <- mutate(brooklyn_shootings, pred = predict(mod))
```

```{r}
summary(mod)
```



```{r}
ggplot(shooting_pred, aes(x = YEAR)) +
  geom_point(aes(y = number_of_crimes, color = "Actual Shootings")) +  # Actual data points
  geom_line(aes(y = pred, color = "Predicted Shootings"), size = 1) +  # Regression line
  labs(title = "Brooklyn Shooting Incidents Over Years",
       x = "Year",
       y = "Number of Shootings") +
  scale_color_manual(values = c("Actual Shootings" = "blue", "Predicted Shootings" = "red")) +
  theme_minimal()
```

## Conclusion and Bias Identification

The analysis of NYPD shooting incidents from 2006 to 2024 highlights key trends and patterns in gun violence across New York City, with a particular focus on Brooklyn due to its consistently high number of reported incidents. Our linear regression model suggests a gradual decline in shooting incidents in Brooklyn, with an estimated reduction of approximately 23 incidents per year. 
I found that the NYPD dataset is racially biased and unfair treatment towards the minority. 